---
Titanic Death Prediction Model
==============================
author: "Peter"

Setup
-----
Set working directory to place with files and load files
  setwd('C:/Users/Pete/titanic-kaggle')
  train <- read.csv("C:/Users/Pete/titanic-kaggle/train.csv")
  View(train)
  test <- read.csv("C:/Users/Pete/titanic-kaggle/test.csv")
  View(test)

First look at Data
------------------
Basic note: test has 418 observations and 11 variables.
train has 891 observations and 12 variables.
The 1 less variable is whether the passengers survived or not.
Create a model using train and then test the model to see who in "test" survived.

```{r}
str(train)
#3 data types here, int, num, and factor. int is integer, num is floating, and
#factor is a category, which can hold strings.
table(train$Survived)
#This outputs a table, row1 is [0,1], row2 is [549, 342]. as in 549 died and 342 survived
prop.table(table(train$Survived))
#Gives proportions, and we see that 61.61% of passengers died while 38.38% survived in the training set
```
1st Hypothesis
--------------
As a first prediction, since the majority of patients died in my train set, let us say everyone died.

```{r}
test$Survived <- rep(0,418)
```
This creates a new column called "Survived" and inputs 0 for 418 rows.
If a "Survived" column already existed, then the values would be replaced.
```{r}
submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
#These words are case sensitive
write.csv(submit,file = "Titanic_Model.csv",row.names = FALSE)
#Create csv file using data frame and exclude row numbers
#Submission to Kaggle requires passengerID and Survived prediction, 2 columns with 418 observations.
```
Results from first submission:

Prediction score of: 0.62679
- - - -
Gender-Based analysis
---------------------
```{r}
summary(train$Sex)
#female 314, male 577
prop.table(table(train$Sex,train$Survived),1)
#The 1 makes it so that I get row-wise proportions, 2 gives column-wise proportions
#74% of females survived, while 81% of males died
```

```{r}
test$Survived <- 0
#Creates a whole column of 0
test$Survived[test$Sex == 'female'] <- 1
#From everyone died, change so that only men died. All women survived.
submit <- data.frame(PassengerId = test$PassengerId, Survived = test$Survived)
write.csv(submit,file = "Titanic_Model.csv",row.names = FALSE)
```

Prediction score of: 0.76555
